Theory of Phonon Polaritons in Polar Dielectrics

### 2018-08-05 paper.tex
I found some repeated sections, commented. Made some other edits. I noticed that equations and figures have explicit callouts rather than using ~\ref and ref labels. Could either Nick or Jenny fix this?

Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function. Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion. As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components. However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$. In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.  Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone. It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons. However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet. Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields. These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.

\section{Application to bulk materials}
% calculation basics
To demonstrate the practical use of this theoretical framework, we present an example calculation of the dielectric function for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in Figure 1(a). 
Phonon bandstructures for SiC and hBN were calculated within density functional perturbation
theory (DFPT) as implemented in the ABINIT package\cite{abinit1,abinit2,abinit3,Gonze1997dynamical,Hamann2005metric}, with electric field response included to appropriately capture LO-TO splitting. Initial ground state density functional theory (DFT) properties were predicted using norm-conserving Vanderbilt pseudopotentials of the PBEsol parameterization for SiC, and PBE with a D2 (Grimme) van der Waals correction for hBN \cite{PBEsol,PBE,ONCV_PPs,pseudodojo,grimme}. We calculated the phonon modes, their frequencies (shown in Figure 1(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation (12), using experimental values for the dissipation rates $\Gamma$. We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering and three-phonon decay processes.

% about the scissor shift -- and a defense sentence in case someone hits us with "semi-empirical" later on
Additionally, for SiC a scissor shift of 1.01 eV was applied to shift the calculated band gap of 2.26 eV to match the experimental gap of $\sim$3.27 eV \cite{band_gap_expt1,band_gap_expt2}. This is necessary to correct for the well-known underestimation of the band gap in ground state density functional theory calculations and the corresponding effects to the dielectric response. In materials like hBN where excitonic effects play a significant role, the optical properties of the system are not determined solely by the energy associated with band to band recombination, and instead must also account for the exciton binding energy\cite{robert2016excitonic}. 
Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{levinshtein2001properties} and E$_{b}$ = 0.67 eV from recent \emph{ab initio} calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV. As the bandgap we compute from DFT is 4.48 eV, we did not require a correction to the gap in hBN.
From these calculations, the high-frequency dielectric tensor for SiC was found to be $\epsilon^{\infty}_{\parallel}$ = 6.986, $\epsilon^{\infty}_{\perp}$ = 7.307 and for hBN, $\epsilon^{\infty}_{\parallel}$ = 4.634, $\epsilon^{\infty}_{\perp}$ = 2.593.  This also leads to a modification of the phonon spectra. % For a material in which the band gap is not as well known as in SiC, a more computationally intensive GW calculation could be used for an entirely \it{ab initio} implementation of our theory. If excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter prediction might be used to further improve predictive accuracy. 
The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figure \ref{fig:epsilon}. As can be seen, the agreement between the theoretical calculation and the experimental results is quite good for silicon carbide, and good for the upper Reststrahlen band of hBN. For the lower Reststrahlen band, the major discrepancy results from a roughly 5\% deviation between calculated and measured transverse optical phonon frequencies.

@@

In future work, besides furnishing an \emph{ab initio} description of nonlocality and reduced-dimension phonon polaritons, there are a number of interesting questions that can be addressed by the framework provided here. For example, we may consider situations such as the phonon polaritons associated with optical phonons at the interface of two materials. Another system of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other. In that case, it would be relevant to see how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.

>>>>

%section below appears twice - commenting it out.

%Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on `nonlocality' in the phonon contribution to the dielectric function. Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion. As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components. However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$. In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.  Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone. It would then appear  to see nonlocal behavior for the phonon-polaritons. However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet. Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields. These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.

\section{Application to bulk semiconductor systems}
% calculation basics
To demonstrate the practical use of this theoretical framework, we present an example calculation of the dielectric function for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in Figure 1(a). 
Phonon bandstructures for SiC and hBN were calculated within density functional perturbation
theory (DFPT) as implemented in the ABINIT package\cite{abinit1,abinit2,abinit3,Gonze1997dynamical,Hamann2005metric}, with electric field response included to appropriately capture LO-TO splitting. Initial ground state density functional theory (DFT) properties were predicted using norm-conserving Vanderbilt pseudopotentials of the PBEsol parameterization for SiC, and PBE with a D2 (Grimme) van der Waals correction for hBN \cite{PBEsol,PBE,ONCV_PPs,pseudodojo,grimme}. We calculated the phonon modes, their frequencies (shown in Figure 1(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation (12), using experimental values for the dissipation rates $\Gamma$. We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering and three-phonon decay processes.

% about the scissor shift -- and a defense sentence in case someone hits us with "semi-empirical" later on
Additionally, for SiC a scissor shift of 1.01 eV was applied to shift the calculated band gap of 2.26 eV to match the experimental gap of $\sim$3.27 eV \cite{band_gap_expt1,band_gap_expt2}. This is necessary to correct for the well-known underestimation of the band gap in ground state density functional theory calculations and the corresponding effects to the dielectric response. In materials like hBN where excitonic effects play a significant role, the optical properties of the system are not determined solely by the energy associated with band to band recombination, and instead must also account for the exciton binding energy\cite{robert2016excitonic}. 
Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{levinshtein2001properties} and E$_{b}$ = 0.67 eV from recent \emph{ab initio} calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV. As the bandgap we compute from DFT is 4.48 eV, we did not require a correction to the gap in hBN.
From these calculations, the high-frequency dielectric tensor for SiC was found to be $\epsilon^{\infty}_{\parallel}$ = 6.986, $\epsilon^{\infty}_{\perp}$ = 7.307 and for hBN, $\epsilon^{\infty}_{\parallel}$ = 4.634, $\epsilon^{\infty}_{\perp}$ = 2.593.  This also leads to a modification of the phonon spectra.
Note that for a materials where the optical band gap is not as well known as in SiC, a more computationally intensive GW calculation could be used while retaining the \emph{ab initio} implementation of our theory. Further, if excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter Equation prediction might be used to further improve predictive accuracy within the same framework presented here.
The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figure \ref{fig:epsilon}. As can be seen, the agreement between the theoretical calculation and the experimental results is quite good for silicon carbide, and good for the upper Reststrahlen band of hBN. For the lower Reststrahlen band, the major discrepancy results from a roughly 5\% deviation between calculated and measured transverse optical phonon frequencies.

@@

In future work, besides furnishing an \emph{ab initio} description of nonlocality and reduced-dimension phonon polaritons, there are a number of interesting questions that can be addressed by the framework provided here. For example, we may consider situations such as the phonon polaritons associated with optical phonons at the interface of two materials. Another system of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other. In that case, it would be relevant to evaluate how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.
